The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-24 script/en/chapters/numbertheory.tex
defined new environment "example"; made vertical space of listoffigures/listoftables and listof{sagecode} the same
Taged remarks ans examples with \begin{remark} etc.(de + en)

{\bf Example:} \\
$$\frac{19}{7} = 2 + \frac{5}{7} \quad (m=7, ~c = 2, ~r = 5)$$

@@

\newpage
{\bf Examples:}
\begin{itemize}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{itemize}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

{\bf Examples:}
\begin{itemize}
   \item[] $19 \equiv 12 {\rm ~(mod~} 7)$,         
           because the remainders are equal:  $19 / 7 = 2$ remainder $5$  and  $12 / 7 = 1$ remainder $5$.
   \item[] $23103 \equiv 0 {\rm ~(mod~} 453)$, because $23103 / 453 = 51$ remainder $0$  and  $0 / 453 = 0$ remainder $0$.
\end{itemize}

@@

{\bf Comment:}\\
The above equivalence does not apply to the sum $(a + b)$!

{\bf Example: }\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$); but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem \ref{thm-zth-div} does not even apply to sums in one direction. It is correct for sums only if the 
remainder is 0 and only in the following direction:  if a divisor divides both summands with no remainder, it also 
divides the sum with no remainder.

@@

{\bf Example:} \\
Is $69,993$ divisible by $7$? \\
The number can be written in the form of a difference in which it is clear that each operand is divisible by $7$: 
$69,993 = 70,000 - 7$. Therefore, the difference is also divisible by $7$.

@@

{\bf Example:}
\begin{eqnarray*}
2 + 4 \equiv 6 \equiv 1 {\rm ~(mod~ } 5) ; \quad 1 + 1 \equiv 2 \equiv 2 {\rm ~(mod~ } 5),  \nonumber \\
2 * 4 \equiv 8 \equiv 3 {\rm ~(mod~ } 5) ; \quad 3 * 4 \equiv 12 \equiv 2 {\rm ~(mod~ } 5). \nonumber
\end{eqnarray*}

@@

This is not the case, however, for 
modular multiplication.
\begin{theorem}\label{thm-zth-multinv}
A natural number $a$ from the set $\{1, \cdots, m-1\}$ has one inverse if
and only if this number and the modulo $m$ are co-prime\footnote{Two whole
numbers $a$ and $b$ are co-prime\index{Number!co-prime} if and only if
${\rm gcd}(a, b) = 1$.\\
If $p$ is prime and $a$ is a random whole number that is not a multiple of
$p$, then $p$ and a are co-prime.\\
Further name to the topic co-prime (with $a_i \in \mathbb{Z}, i=1, \cdots, n$):
\begin{enumerate}
\item $a_1,a_2, \cdots, a_n$ are {\em relatively prime} \index{Relatively prime}, if $ \mbox{gcd}(a_1, \cdots , a_n) =1.$
\item An even stronger request for more than two numbers is :\\
                $a_1, \cdots , a_n$ are {\em in pairs relatively prime}, if for all $i=1, \cdots, n$ and 
$j=1, \cdots , n$ with $ i \neq j $: $ \mbox{gcd} (a_i, a_j) =1. $
\end{enumerate}
Example: $2,3,6 $ are relatively prime, because $ \mbox{gcd} (2,3,6)=1.$ 
They are not in pairs relatively prime, because $ \mbox{gcd} (2,6)=2>1.$}\index{Gcd}, in other words if $a$ and $m$ have 
no common prime factors.
\end{theorem}
Since $m=5$ is prime, the numbers $1$ to $4$ are relatively prime to $5$ and {\bf each} of these numbers has a 
multiplicative inverse in mod $5$.

@@

{\bf Examples:}
\begin{itemize}
\item[a)] In $\mathbb{Z}_{15}$, for example, the element $5$ does not have an inverse.
          That is to say, there is no $a$ with \\ $5 * a \equiv 1 {\rm ~(mod~}15).$
          Each modulo product with $5$ on this set gives $5, 10$ or $0$.
\item[b)] In $\mathbb{Z}_{55} \setminus \{0\}$, for example, the elements $5$ and $11$ do not have multiplicative inverses.
          That is to say, there is no $a \in \mathbb{Z}_{55}$ such that $5 * a\equiv 1~(~mod~55~) $ and no $a$ such that $11*a \equiv 1~(~mod~55~)$. 
          This is because $5$ and $11$ are not relatively prime to $55$.
          Each modulo product with $5$ on this set gives $5, 10, 15, \dots, 50$ or $0$.
          Each modulo product with $11$ on this set gives $11, 22, 33, 44$ or $0$.
\end{itemize}
On the other hand, there are subsets of $\mathbb{Z}_n$ that form a group with the operation mod*.
If we choose all elements in $\mathbb{Z}_n$ that are relatively prime to $n$, then this set forms a group with the operation mod*.
We call this set $\mathbb{Z}_n^*$.

@@

{\bf Example:}
For $n=10=2*5$ the following applies: 
\begin{itemize}
  \item[] \index{Remainder set!full} full remainder set $R = \mathbb{Z}_n = \{ 0, 1, 2, 3, 4, 5, 6, 7, 8, 9 \}$
  \item[] \index{Remainder set!reduced} reduced remainder set $R' = \mathbb{Z}_n^* = \{ 1, 3, 7, 9 \} \longrightarrow J(n)=4$.
\end{itemize}

{\bf Comment:}
$R'$ or $\mathbb{Z}_n^*$ is always a genuine subset of $R$ or $\mathbb{Z}_n$ because $0$ is always an element of $R$ but never 
an element of $R'$. Since $1$ and $n-1$ are always relatively prime to $n$, they are always elements of both sets.

@@

Example: For $a = 3$ modulo $10$ with $\mathbb{Z}_n^* = \{ 1, 3, 7, 9 \}$ it holds that $a^{-1} = 7$:
\begin{eqnarray*}
3 & \equiv & 3 * 1{\rm ~(mod~}10), \nonumber \\
9 & \equiv & 3 * 3{\rm ~(mod~}10), \nonumber \\
1 & \equiv & 3 * 7{\rm ~(mod~}10), \nonumber \\
7 & \equiv & 3 * 9{\rm ~(mod~}10). \nonumber 
\end{eqnarray*}

@@

{\bf Examples:} 
\begin{itemize}
\item  $n=70=2*5*7 \Longrightarrow $ using theorem \ref{J_of_p1..pk}: $ J(n)= 1\cdot 4 \cdot 6 =24.$
\item  $n=9=3^2 \Longrightarrow$ using theorem \ref{J_of_n}: $ J(n)= 3^1\cdot 2 =6,$ because  $\mathbb{Z}_9^* =\{ 1,2,4,5,7,8\}.$
\item $n = 2,701,125 = 3^2 * 5^3 * 7^4 \Longrightarrow $ using theorem \ref{J_of_n}: 
$$J(n) = [3^1 * 2] * [5^2 * 4] * [7^3 * 6] = 1,234,800.$$
\end{itemize}

@@

{\bf Example:}\par
Find the multiplicative inverse of $1579$ modulo $7351$.\\
According to Euler-Fermat:  $a^{J(n)} = 1 {\rm ~(mod~} n)$ for all $a$ in $\mathbb{Z}_n^*$.
If we divide both sides by $a$, we get: $a^{J(n) - 1} \equiv a^{-1} {\rm ~(mod~} n)$.
For the special case that the modulo is prime, we have $J(n) = p - 1$.
Therefore, the modular inverse is 
$$a^{-1} = a^{J(n) - 1} \equiv a^{(p-1)-1} \equiv a^{p-2} {\rm ~(mod~} p).$$
For our example, this means:
\begin{itemize}
\item[] Since the modulus $7351$ is prime, $p-2 = 7349$. \\
    $1579^{-1} \equiv 1579^{7349} {\rm ~(mod~} p).$
\end{itemize}
By cleverly breaking down the exponent, we can calculate this power relatively easily (see Section~\ref{hohpot}
\hyperlink{hohpot}{Fast calculation of high powers}):
\begin{itemize}
\item[] $7349 = 4096 + 2048 + 1024 + 128 + 32 + 16 + 4 + 1$
\item[] $1579^{-1} \equiv 4716 {\rm ~(mod~} 7351).$
\end{itemize}

@@

\newpage
The following two tables show the multiplicative orders and primitive roots modulo $45$ and modulo $46$.

@@

% --------------------------------------------------------------------------
\hypertarget{RSA}{}
\subsection{How the RSA procedure works}
\label{RSA}
The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem \ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source
Initiative\index{Open Source}, that one can claim patents on algorithms.
But there is really no understanding outside the U.S., that one can get
a patent for a special case (3 factors) of an algorithm (RSA), although
the patent for the general case was almost expired.}$^,$\footnote{If the
two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is
not true for all $ m<n$ (although $ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
Example: If $ n=5^2$ then according to theorem \ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$ 
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

@@

{\bf Comment:}\par

The security of the RSA algorithm depends as with all public key methods on the difficulty to calculate the private
key $d$ from the public key $(n,e)$.

@@

{\bf Comment:} \\
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice)\index{Mathematica}:
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\index{Pari-GP}\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or \index{BC} BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\index{Mathematica}\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.\index{Pari-GP}
\vskip +6 pt  % \vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +18 pt
} is extremely difficult to solve.

@@

{\bf Example:} \\
Using the extended gcd we can determine for $e = 37$ the multiplicative inverse number $d$ to modulo $3588$ (i.e.
$37*d \equiv 1 {\rm ~(mod~} 3588$)):

{\tt 0.}
 $ \left(\begin{array}{cccc} 3588 & | & 1 & 0 \\ 37 & | & 0 & 1 \end{array} \right)$

>>>>

\begin{example}
$$\frac{19}{7} = 2 + \frac{5}{7} \quad (m=7, ~c = 2, ~r = 5)$$
\end{example}

@@

\newpage
\begin{example}
\begin{itemize}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{itemize}
\end{example}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

\begin{example}
\begin{itemize}
   \item[] $19 \equiv 12 {\rm ~(mod~} 7)$,         
           because the remainders are equal:  $19 / 7 = 2$ remainder $5$  and  $12 / 7 = 1$ remainder $5$.
   \item[] $23103 \equiv 0 {\rm ~(mod~} 453)$, because $23103 / 453 = 51$ remainder $0$  and  $0 / 453 = 0$ remainder $0$.
\end{itemize}
\end{example}

@@

\begin{remark}
The above equivalence does not apply to the sum $(a + b)$!
\end{remark}

\begin{example}
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$); but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem \ref{thm-zth-div} does not even apply to sums in one direction. It is correct for sums only if the 
remainder is 0 and only in the following direction:  if a divisor divides both summands with no remainder, it also 
divides the sum with no remainder.
\end{example}

@@

\begin{example}
Is $69,993$ divisible by $7$? \\
The number can be written in the form of a difference in which it is clear that each operand is divisible by $7$: 
$69,993 = 70,000 - 7$. Therefore, the difference is also divisible by $7$.
\end{example}

@@

\begin{example}
\begin{eqnarray*}
2 + 4 \equiv 6 \equiv 1 {\rm ~(mod~ } 5) ; \quad 1 + 1 \equiv 2 \equiv 2 {\rm ~(mod~ } 5),  \nonumber \\
2 * 4 \equiv 8 \equiv 3 {\rm ~(mod~ } 5) ; \quad 3 * 4 \equiv 12 \equiv 2 {\rm ~(mod~ } 5). \nonumber
\end{eqnarray*}
\end{example}

@@

This is not the case, however, for 
modular multiplication.
\begin{theorem}\label{thm-zth-multinv}
A natural number $a$ from the set $\{1, \cdots, m-1\}$ has one inverse if
and only if this number and the modulo $m$ are co-prime\footnote{Two whole
numbers $a$ and $b$ are co-prime\index{Number!co-prime} if and only if
${\rm gcd}(a, b) = 1$.\\
If $p$ is prime and $a$ is a random whole number that is not a multiple of
$p$, then $p$ and a are co-prime.\\
Further name to the topic co-prime (with $a_i \in \mathbb{Z}, i=1, \cdots, n$):
\begin{enumerate}
\item $a_1,a_2, \cdots, a_n$ are {\em relatively prime} \index{Relatively prime}, if $ \mbox{gcd}(a_1, \cdots , a_n) =1.$
\item An even stronger request for more than two numbers is :\\
                $a_1, \cdots , a_n$ are {\em in pairs relatively prime}, if for all $i=1, \cdots, n$ and 
$j=1, \cdots , n$ with $ i \neq j $: $ \mbox{gcd} (a_i, a_j) =1. $
\end{enumerate}
\begin{example} $2,3,6 $ are relatively prime, because $ \mbox{gcd} (2,3,6)=1.$
\end{example}
They are not in pairs relatively prime, because $ \mbox{gcd} (2,6)=2>1.$}\index{Gcd}, in other words if $a$ and $m$ have 
no common prime factors.
\end{theorem}
Since $m=5$ is prime, the numbers $1$ to $4$ are relatively prime to $5$ and {\bf each} of these numbers has a 
multiplicative inverse in mod $5$.

@@

\begin{example}
\begin{itemize}
\item[a)] In $\mathbb{Z}_{15}$, for example, the element $5$ does not have an inverse.
          That is to say, there is no $a$ with \\ $5 * a \equiv 1 {\rm ~(mod~}15).$
          Each modulo product with $5$ on this set gives $5, 10$ or $0$.
\item[b)] In $\mathbb{Z}_{55} \setminus \{0\}$, for example, the elements $5$ and $11$ do not have multiplicative inverses.
          That is to say, there is no $a \in \mathbb{Z}_{55}$ such that $5 * a\equiv 1~(~mod~55~) $ and no $a$ such that $11*a \equiv 1~(~mod~55~)$. 
          This is because $5$ and $11$ are not relatively prime to $55$.
          Each modulo product with $5$ on this set gives $5, 10, 15, \dots, 50$ or $0$.
          Each modulo product with $11$ on this set gives $11, 22, 33, 44$ or $0$.
\end{itemize}
\end{example}
On the other hand, there are subsets of $\mathbb{Z}_n$ that form a group with the operation mod*.
If we choose all elements in $\mathbb{Z}_n$ that are relatively prime to $n$, then this set forms a group with the operation mod*.
We call this set $\mathbb{Z}_n^*$.

@@

\begin{example}
For $n=10=2*5$ the following applies:
\end{example}
\begin{itemize}
  \item[] \index{Remainder set!full} full remainder set $R = \mathbb{Z}_n = \{ 0, 1, 2, 3, 4, 5, 6, 7, 8, 9 \}$
  \item[] \index{Remainder set!reduced} reduced remainder set $R' = \mathbb{Z}_n^* = \{ 1, 3, 7, 9 \} \longrightarrow J(n)=4$.
\end{itemize}

\begin{remark}
$R'$ or $\mathbb{Z}_n^*$ is always a genuine subset of $R$ or $\mathbb{Z}_n$ because $0$ is always an element of $R$ but never 
an element of $R'$. Since $1$ and $n-1$ are always relatively prime to $n$, they are always elements of both sets.
\end{remark}

@@

\begin{example} $a = 3$ modulo $10$ with $\mathbb{Z}_n^* = \{ 1, 3, 7, 9 \}$ it holds that $a^{-1} = 7$:
\end{example}
\begin{eqnarray*}
3 & \equiv & 3 * 1{\rm ~(mod~}10), \nonumber \\
9 & \equiv & 3 * 3{\rm ~(mod~}10), \nonumber \\
1 & \equiv & 3 * 7{\rm ~(mod~}10), \nonumber \\
7 & \equiv & 3 * 9{\rm ~(mod~}10). \nonumber 
\end{eqnarray*}

@@

\begin{example} 
\begin{itemize}
\item  $n=70=2*5*7 \Longrightarrow $ using theorem \ref{J_of_p1..pk}: $ J(n)= 1\cdot 4 \cdot 6 =24.$
\item  $n=9=3^2 \Longrightarrow$ using theorem \ref{J_of_n}: $ J(n)= 3^1\cdot 2 =6,$ because  $\mathbb{Z}_9^* =\{ 1,2,4,5,7,8\}.$
\item $n = 2,701,125 = 3^2 * 5^3 * 7^4 \Longrightarrow $ using theorem \ref{J_of_n}: 
$$J(n) = [3^1 * 2] * [5^2 * 4] * [7^3 * 6] = 1,234,800.$$
\end{itemize}
\end{example}

@@

\begin{example}
\par
Find the multiplicative inverse of $1579$ modulo $7351$.\\
\end{example}
According to Euler-Fermat:  $a^{J(n)} = 1 {\rm ~(mod~} n)$ for all $a$ in $\mathbb{Z}_n^*$.
If we divide both sides by $a$, we get: $a^{J(n) - 1} \equiv a^{-1} {\rm ~(mod~} n)$.
For the special case that the modulo is prime, we have $J(n) = p - 1$.
Therefore, the modular inverse is 
$$a^{-1} = a^{J(n) - 1} \equiv a^{(p-1)-1} \equiv a^{p-2} {\rm ~(mod~} p).$$
For our example, this means:
\begin{itemize}
\item[] Since the modulus $7351$ is prime, $p-2 = 7349$. \\
    $1579^{-1} \equiv 1579^{7349} {\rm ~(mod~} p).$
\end{itemize}
By cleverly breaking down the exponent, we can calculate this power relatively easily (see Section~\ref{hohpot}
\hyperlink{hohpot}{Fast calculation of high powers}):
\begin{itemize}
\item[] $7349 = 4096 + 2048 + 1024 + 128 + 32 + 16 + 4 + 1$
\item[] $1579^{-1} \equiv 4716 {\rm ~(mod~} 7351).$
\end{itemize}

@@

The following two tables show the multiplicative orders and primitive roots modulo $45$ and modulo $46$.

@@

% --------------------------------------------------------------------------
\hypertarget{RSA}{}
\subsection{How the RSA procedure works}
\label{RSA}
The individual steps for implementing the \index{RSA} RSA procedure can be described as follows (see \cite[p. 213 ff]{Eckert2003}\index{Eckert 2003} and 
\cite[p. 338 ff]{Sedgewick1990}\index{Sedgewick 1990}). 
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption, and steps 
6 and 7 are the decryption:
\begin{itemize}
\item[{\bf 1.}] Select two distinct random prime numbers\footnote{
Compaq introduced the so-called multi-prime method with high marketing 
effort in 2000.
$n$ was the product of two big and one relative small prime: $n=o*p*q.$ 
With theorem \ref{J_of_p1..pk} we get:
$J(n)= (o-1)*(p-1)*(q-1). $ This method did not assert itself yet.\\
One reason probably is, that Compaq claimed a patent \index{Patent} on it. 
Generally there is less understanding in Europe and with the Open Source
Initiative\index{Open Source}, that one can claim patents on algorithms.
But there is really no understanding outside the U.S., that one can get
a patent for a special case (3 factors) of an algorithm (RSA), although
the patent for the general case was almost expired.}$^,$\footnote{If the
two primes $p$ and $q$ are equal then $ (m^{e})^{d} \equiv m $ mod $n$ is
not true for all $ m<n$ (although $ e*d \equiv 1$ mod $J(n)$ is fulfilled). 
\begin{example} If $ n=5^2$ then according to theorem \ref{thm-zth-phinum} it 
is $J(n)=5*4=20, ~ e=3, ~ d=7, ~e*d=21\equiv 1$ mod $J(n).$
\end{example}
But it is $ (5^3)^7\equiv 0 $ mod $ 25$. } $p$ and $q$ and calculate 
$n = p*q$\footnote{The GISA\index{GISA} (German Information Security Agency) 
recommends, to choose the prime factors $p$ and $q$ almost the same,
but not too close:
$$ 0.5 < |\log_2 (p) - \log_2 (q) | <30. $$
They recommend to generate the primes independently and check that the 
restriction is fulfilled (see \cite{GISA2002}).
}.\\
The value $n$ is called the RSA modulus\footnote{
In CrypTool\index{CrypTool} the RSA modulo is denoted with a capital ``$N$'' .}.

@@

\begin{remark}\par

The security of the RSA algorithm depends as with all public key methods on the difficulty to calculate the private
key $d$ from the public key $(n,e)$.
\end{remark}

@@

\begin{remark}
In this example using such small numbers, it would be possible, but with large numbers the discrete logarithm\index{Discrete logarithm}
problem\footnote{If you try to determine the discrete logarithm $x$ that solves the equation $11^x \equiv 49 {\rm ~(mod~} 347) $ with 
Mathematica by means of Solve, you obtain the {em tdep message} ``The equations appear to involve the variables to be solved for 
in an essentially non-algebraic way''. Mathematica therefore claims not to know a direct algebraic procedure for solving the equation.
Yet Mathematica is able to calculate this with the general function for the multiplicative order (here for Alice)\index{Mathematica}:
{\tt MultiplicativeOrder[11, 347, 49]} delivers the value $67$.\\
The syntax with Pari-GP is: \texttt{znlog(Mod(49,347),Mod(11,347))}.\index{Pari-GP}\\
Such number-theory tasks can also be solved using other tools such as the
LiDIA\index{LiDIA} or \index{BC} BC package (see web links in appendix).
The dl function in the {\bf LC} user interface for LiDIA also delivers the value $67$ for {\tt dl(11,49,347)}.
}${}^,$\footnote{Why have the functions delivered the value $67$ rather than $240$ for the dl problem for Alice?
The discrete logarithm is the smallest natural exponent that solves the equation $11^x\equiv49 {\rm ~(mod~} 347)$. Both $x=67$ 
and $x=240$ (the number selected in the example) satisfy the equation and can therefore be used to calculate the session key: 
$285^{240} \equiv 285^{67}\equiv 268 {\rm ~(mod~} 347)$.
If Alice and Bob had selected a primitive root modulo $p$ as base $g$, then for every remainder from the set 
$\{1, 2, \cdots, p-1\}$ there is exactly one exponent from the set $\{0, 1, \cdots, p-2\}$.\\
For info: there are 
$172$ different primitive roots for modulo $347$, $32$ of which are prime (not necessary).
Since the number $11$ selected for $g$ in the example is not a primitive root of $347$, the remainders do not take all 
values from the set $\{1, 2, \cdots, 346\}$. Thus, for a particular remainder there may be more than one exponent or 
even no exponent at all in the set $\{0, 1, \cdots, 345\}$ that satisfies the equation.\index{Mathematica}\\
{\tt 
PrimeQ[347] = True; EulerPhi[347] = 346; GCD[11, 347] = 1;
MultiplicativeOrder[11, 347] = 173}\\
The syntax with Pari-GP is: \texttt{isprime(347); eulerphi(347);
  gcd(11,347); znorder(Mod(11,347))}.\index{Pari-GP}
\vskip +6 pt  % \vskip +10 pt
\begin{tabular}{|c|c|l|}
\hline
i  & $11^i{\rm ~mod~}347$ & \\
\hline
      0  &          1   &  \\
      1  &         11   &  \\                                     
      2  &        121   &  \\                                     
      3  &        290   &  \\                                     
     67  &         49   & searched exponent \\                    
    172  &        284   &  \\                                                  
    173  &          1   &= multiplicative order of $11 {\rm  ~(mod~} 347)$ \\ 
    174  &         11   &  \\                                                     
    175  &        121   &  \\                                     
    176  &        290   &  \\                                     
    240  &         49   & searched exponent \\
\hline
\end{tabular}
\vskip +18 pt
} is extremely difficult to solve.
\end{remark}

@@

\begin{example}
Using the extended gcd we can determine for $e = 37$ the multiplicative inverse number $d$ to modulo $3588$ (i.e.
$37*d \equiv 1 {\rm ~(mod~} 3588$)): 
\end{example}
{\tt 0.}
 $ \left(\begin{array}{cccc} 3588 & | & 1 & 0 \\ 37 & | & 0 & 1 \end{array} \right)$
